---
title: "R FINAL Assignment"
author: "J"
output: pdf_document
---

# Load Packages 
```{r}
library(haven) #read STATA Files
library(dplyr) #data wrangling 
library(janitor) #clean names for variables
library(emmeans) #graphs and plots
library(ggplot2) #data visualization
```

#Import Dataset and load the dataset with clean variable names and convert Stata missing values
```{r}
nsch <- read_dta("~/Desktop/data/nsch_2023e_topical_Stata/nsch_2023e_topical.dta") %>%
  clean_names()

dim(nsch)
head(nsch)

nsch <- haven::read_dta("~/Desktop/data/nsch_2023e_topical_Stata/nsch_2023e_topical.dta") %>%
  dplyr::mutate(across(everything(), haven::zap_missing))
```

#Select variables for analysis which include insurance coverage, poverty ratio, language, metropolitan area, learning disability, race, and sex
```{r}
nsch_small <- nsch %>%
  select(currins, fpl_i1, hhlanguage, k2q30a,k2q30c, k2q34b,k2q34c,k4q28x04, k4q36, metro_yn, sc_race_r, sc_sex)
```

#Recode categorical variables and make it more readable 
```{r}
nsch_small <- nsch %>%
  mutate(
    #Insurance
    curr_ins = factor(currins, 
                      levels = c(1,2),
                      labels = c("Insured", "Not Insured")),
    #Poverty ratio
    poverty_group = cut(fpl_i1,
                        breaks = c(0,100,200,400,Inf), 
                        labels = c("0-99% FPL", "100-199 FPL", "200-399% FPL","400%+FPL"),
                        right = FALSE),
    #Language spoken 
    hh_language = factor(hhlanguage,
                         levels = c(1,2,3),
                         labels = c("English", "Spanish", "Other")),
    
    #Learning disability 
    learn_dis = factor(k2q30a,
                       levels = c(1,2),
                       labels = c("Yes", "No")),
   
    #Care not received 
    
    care_unmet = factor(k4q28x04, 
                        levels = c(1,2),
                        labels = c("Yes", "No")),
    
    #race 
    
    race = factor(sc_race_r, 
                  levels = c(1,2,3,4,5,7),
                  labels = c("White alone", "Black or African American alone", "American Indian or Alaska Native alone", "Asian alone", "Native Hawaiian and Other Pacific Islander alone", "Two or More Races")),
    
    #metropolitan status
    
    metro = factor(metro_yn, 
                   levels =c(1,2),
                   labels = c("MSA", "Non-MSA")),
    
    #Sex 
    sex = factor(sc_sex,
                 levels = c(1,2),
                 labels = c("Female", "Male"))
                        
)
```

#Filtering data, which includes excluding rows with missing insurance, race, sex, or disability, and dropping "Other" household language 
#Summarize the proportion uninsured and the proportion with a learning disability 
#each subgroup is defined by race and subgroup status

``` {r}
nsch_summary <- nsch_small %>%
  filter(!is.na(curr_ins),
         !is.na(race),
         !is.na(sex),
         !is.na(learn_dis),
         hh_language %in% c("English","Spanish")) %>%
  mutate(
    uninsured_dummy = ifelse(curr_ins == "Not Insured", 1, 0),
    disability_dummy = ifelse(learn_dis == "Yes", 1, 0)
    )%>%
  group_by(race, metro, sex) %>%
  summarize(
    prop_uninsured = mean(uninsured_dummy, na.rm = TRUE),
    prop_disabled = mean(disability_dummy, na.rm = TRUE),
    n = n()
)
print(nsch_summary)
```

The analysis shows differences in health coverage and disability status across race and metropolitan status. For example, uninsured rates are higher among black and American Indian or Alaskan natives. Within groups, there are also differences with blacks in non-MSA areas, having a greater proportion of uninsured people and greater learning disability rates. These findings are consistent with existing literature and research that children from minority backgrounds and those in non-urban areas face more barriers to health insurance coverage and healthcare access. 

```{r}
# 1. Race and Learning Disability - bivariate analysis tests
# Create contingency table
table_race <- table(nsch_small$learn_dis, nsch_small$race)

# Run chi-square test
chisq.test(table_race)

#proportions by race
prop.table(table_race, margin = 2)  # column proportions

# 2. Sex and Learning Disability
table_sex <- table(nsch_small$learn_dis, nsch_small$sex)
chisq.test(table_sex)
prop.table(table_sex, margin = 2)

```
```{r}
#Visualizing race and sex differences across children with learning disability
race_sex_summary <- nsch_small %>%
  filter(!is.na(learn_dis), !is.na(race), !is.na(sex)) %>%
  group_by(race, sex) %>%
  summarise(prop_disabled = mean(learn_dis == "Yes"), .groups = "drop")

ggplot(race_sex_summary, aes(x = race, y = prop_disabled, fill = sex)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Learning Disability Identification by Race and Sex",
       x = "Race",
       y = "Percent with Learning Disability",
       fill = "Sex") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

#Fitting Logistic Regression Models
```{r}
nsch_summary2 <- nsch_small %>%
  filter(!is.na(curr_ins),
         !is.na(race),
         !is.na(sex),
         !is.na(learn_dis),
         hh_language %in% c("English","Spanish")) %>%
  mutate(
    uninsured_dummy = ifelse(curr_ins == "Not Insured", 1, 0),
    disability_dummy = ifelse(learn_dis == "Yes", 1, 0)
    )
```

```{r}
model1 <- glm(uninsured_dummy ~ race, 
              data = nsch_summary2, 
              family = "binomial")
summary(model1) 

model2 <- glm(uninsured_dummy ~ metro,
              data = nsch_summary2, 
              family = "binomial")
summary(model2)
```
1) #Identify key numbers for model1 
Estimate for black or african american alone = 0.652 | p value = <2e-16 | odds ratio = 1.91 
American Indian or Alaska native = 1.29 | p value = <2e - 16 | odds ratio = 3.63
Asian alone = -0.08 | p value = 0.47 | odds ratio = 0.92 
Native Hawaiian or Pacific Islander = 1.28 | p value = 5.13 e - 10 |odds ratio = 3.59
Two or more races = -0.03 | p value = 0.70 | odds ratio = 0.97 
Baseline = white children insured
Interpretation
Results of the logistic regression indicated a significant overall association between race and health insurance coverage 
Black or African American children had 1.92 times higher odds of being uninsured than white children, with a statistically significant p-value < 0.001 
American Indian or Native Alaskan children had 3.63 times higher odds of being uninsured, with a statistically significant p-value < 0.001 
Asian children had slightly lower odds of being uninsured (0.92); however, this difference is not significant at p = 0.47 
For Native Hawaiian or Pacific Islanders, there were 3.59 times higher odds of being uninsured as compared to white children, with a statistically significant p-value < 0.001 
For children of two or more races, there were no significant differences at p = 0.7
Model fit statistics show that adding race as a predictor reduced the deviance from 17,375.11 (null model) to 17,230.48, representing a statistically significant improvement in model fit.
The model’s AIC = 17,242.48, suggesting that while race improves explanatory power compared to a model with no predictors, the overall improvement is modest.




2) #Identify key numbers for model2
Estimate for metroNon-MSA = 0.2756
pvalue = 1.41e-06(highly significant, p <0.001)
odss ratio = exp(0.27567) = 1.3
Baseline: Living in metropolitan areas 

Interpretation 
The models show a significant relationship between metro status and being uninsured, as the coefficient is positive: households in non-MSA areas are more likely to be uninsured than those in MSA areas. The results of the logistic regression indicate a positive association between metro status and insurance coverage, as p < 0.001. The odds of a child being uninsured are 1.3 times higher for households in non-MSA areas than for households in MSA areas, indicating that living outside the metropolitan area is associated with a greater likelihood of being uninsured. 
Model fit statistics showed that adding metro status as a predictor reduced the deviance from 15,711.16 (null model) to 15,689.00, with an AIC of 15,692.98, suggesting a small but meaningful improvement in model fit.

```{r}

# Multiple logistic regression
model_multi <- glm(learn_dis ~ race + metro + hhlanguage + curr_ins + sex + poverty_group +care_unmet,
                   data = nsch_small,
                   family = "binomial")

summary(model_multi)

# Odds ratios
exp(coef(model_multi))
exp(confint(model_multi))
```
```{r}
# Interaction model looking at race * sex 
m1 <- glm(learn_dis ~ race * sex + poverty_group + care_unmet + curr_ins,
          data = nsch_small,
          family = "binomial")

# Get coefficients
coef(m1)              # raw log-odds (numeric)
exp(coef(m1))         # odds ratios

# Confidence intervals
exp(confint(m1))
```
```{r}
# Packages
library(emmeans)
library(dplyr)
library(ggplot2)
library(scales)

# --- 1)interaction model ---

m <- m1

# ensure sex order
nsch_small$sex <- factor(nsch_small$sex, levels = c("Female","Male"))

# covariate values that are constant
at_vals <- list(
  poverty_group = "200-399% FPL",
  care_unmet    = "No",
  curr_ins      = "Insured"
)

# Get predicted probabilities (on response scale) by Race x Sex
emm <- emmeans(m, ~ sex | race, type = "response", at = at_vals)
pred_df <- as.data.frame(emm) %>% 
  rename(pred = prob, lcl = asymp.LCL, ucl = asymp.UCL)

# --- 2) Plot 
ggplot(pred_df, aes(x = sex, y = pred, group = race, color = race)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  geom_errorbar(aes(ymin = lcl, ymax = ucl), width = 0.08, alpha = 0.7) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1), limits = c(0, NA)) +
  labs(
    title = "Predicted Probability of Learning Disability",
    subtitle = "Interaction of Race × Sex (holding poverty=200–399% FPL, care unmet=No, insured)",
    x = NULL, y = "Predicted probability", color = "Race"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```
